Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-09-24 paper/numresults.tex
add krylov-aware numerical results

@@

\subsection{Trace of parameter-dependent matrix functions}
\label{subsec:function-trace}

\todo{Comparison plots with other estimators (Krylov-Aware)?}

>>>>

\subsection{Comparison to Krylov-aware stochastic trace estimation}
\label{subsec:krylov-aware}

The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} also works well for the approximation of the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$. It also samples two standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, of which $\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a Krylov-subspace of $\mtx{A}$ and $\mtx{\Psi}$ to estimate the trace of the approximation residual.

We run an optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} in multiple configurations on the example from \refsec{subsec:hamiltonian} and plot their approximation errors for various values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}. For reference, we add the error of the Chebyshev-Nyström++ method on parameters which lead to a comparable run-time.

\begin{figure}[ht]
    \begin{minipage}[c]{.45\linewidth}
        \centering
        \input{plots/krylov_aware_density.pgf}
    \end{minipage}\hfill%
    \begin{minipage}[c]{.45\linewidth}
        %\label{tab:krylov-aware-density}
        \renewcommand{\arraystretch}{1.2}
        \begin{tabular}{@{}lccccc@{}}
        \toprule
        & $b$ & $q$ & $m$ & $n$ & $t$ (s)\\
        \midrule
        KA I & 10 & 30 & 60 & 20 & 9.13 \\
        KA II & 20 & 30 & 30 & 20 & 9.13 \\
        KA III & 10 & 60 & 30 & 40 & 9.13 \\
        KA IV & 10 & 30 & 30 & 20 & 9.13 \\
        CN++ & 10 & - & - & 20 & 9.13 \\
        \bottomrule
        \end{tabular}
        %\caption{Comparison with Krylov-Aware stochastic trace estimation.}
    \end{minipage}
    \caption{Comparison with Krylov-Aware stochastic trace estimation.}
    \label{fig:krylov-aware-density}
\end{figure}

To demonstrate that the Chebyshev-Nyström++ estimator is also effective on problems unrelated to spectral density estimation, we consider the problem of approximating $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain \cite{chen-2023-krylovaware-stochastic}. We compute the pointwise error from the analytic solution as a function of the temperature $\beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.

\begin{figure}[ht]
    \centering
    \input{plots/krylov_aware_spin.pgf}
    \caption{Comparison with Krylov-Aware stochastic trace estimation.}
    \label{fig:krylov-aware-spin}
\end{figure}

Unlike in the spectral density example from \refsec{subsec:hamiltonian}, the numerical rank of the matrix we approximate wildly changes with the parameter $\beta$. Since the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$. \todo{As a consequence, ...}

@@
